% Copyright (c) 2017 Peter A. Audano III
% GNU Free Documentation License Version 1.3 or later
% See the file COPYING.DOC for copying conditions.

\section{Introduction}
\label{sec.intro}

\subsection{About This Document}
\label{sec.intro.aboutdoc}

This document is a work-in-progress. In its current form, it is incomplete and unpolished. We apologize for any inconvenience this causes, and we are actively working to get it up to our standards.


\subsection{About Kestrel}
\label{sec.intro.aboutkestrel}

Kestrel is a first-in-class variant calling toolkit that uses k-mer frequencies to detect regions of variation and resolve variants. Standard approaches require mapping sequence reads to a reference genome and searching for reads that differ from the reference. When alignments fail, sequence reads are often assembled or variants are called directly from a De~bruijn graph, which is computationally expensive. With similar resource requirements as the alignment approach, Kestrel can resolve variation in where alignments fail.

Sequence reads are first decomposed into k-mers, which are short overlapping fragments of the same length. With features new in KAnalyze~\cite{Audano2014} version 2.0.0, Kestrel keeps an on-disk database of frequencies that can be rapidly searched with minimal memory requirements. The reference sequence is also decomposed into a set of k-mers and left in order. With novel algorithms to simplify searching k-mer space, the k-mer database is used to detect and resolve regions of variation.

Kestrel can identify both single nucleotide polymorphism (SNP) and insertion/deletion (indel) variants using evidence in k-mer frequencies. To our knowledge, no k-mer approach has ever been shown to efficiently recover variation over dense SNPs or large indels. These regions of variation may be larger than k-mers and the sequence reads they came from. Using k-mers on the flanks of these regions, the Kestrel algorithm uses the k-mer database to extend into the altered sequence and perform a targeted local assembly guided by a modified Smith~Waterman algorithm. Variant calls are then extracted from the aligned assembly.

Real sequence data is not uniform, and Kestrel has a set of heuristics to deal with it. There are many parameters that can be tuned to control how large variant regions may be and how much computing resources it may use.

For full details on the Kestrel algorithm and implementation, see our bioaRxiv pre-print publication (https://doi.org/10.1101/153619). A peer-reviewed publication is in progress.
